Devito for large scale elastic modelling and anisotropic inversion

### 2020-04-20 paper.tex
fixing layout ...

\usepackage{listings}
% Define slightly more reasonable Listings defaults
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    prebreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
    frame=lines,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    keywordstyle=\color[gray]{0.4}\bfseries,
    commentstyle=\color[gray]{0.65}\itshape,
    numbers=left,
    captionpos=b,
}
\usepackage{color}
\usepackage{fancyvrb}
\usepackage{fvextra}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, breaklines=true}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}

@@

It is with these specifications in mind that Devito, a symbolic domain
specific language (DSL) and compiler for the automatic generation of
finite-difference stencils, has been developed. Originally deigned to
accelerate research and development in exploration geophysics, the
high-level interface, previously described in detail in
\cite{devito-api}, is built on top of \texttt{SymPy} \cite{sympy} and
is inspired by the underlying mathematical formulations and other
high-level DSLs such as FEniCS \cite{fenics} and Firedrake
\cite{firedrake}. This interface allows the user to formulate
wave-equations, and more generally time-dependent PDEs in a simple and
mathematically coherent way. The symbolic API then automatically
generates finite-difference stencils from these mathematical
expressions. One of the main advantages of
\href{https://github.com/devitocodes/devito}{Devito} over other
finite-difference DSLs is that generic expressions such as sparse
operations (i.e.~point source injection or localized measurements) are
fully supported and expressible in a high-level fashion. The second
component of \href{https://github.com/devitocodes/devito}{Devito} is its
compiler (c.f \cite{devito-compiler}) that generates highly optimized C
code. Generated code is then compiled at runtime for the hardware at
hand.

Previous work focused on the DSL and compiler to highlight the potential
application and use cases of Devito. Here, we present a series of
extensions and applications to large-scale three-dimensional problem
sizes as encountered in exploration geophysics. These experiments are
carried out in Cloud-based HPC systems and include elastic forward
modelling using distributed-memory parallelism and imaging based on the
tilted transverse isotropic (TTI) wave-equation
(\cite{virieux, thomsen1986, zhang-tti, duveneck, louboutin2018segeow}).
These proof of concepts highlight two critical features: First, the
ability of the symbolic interface and the compiler to translate to
large-scale adjoint-based inversion problems that require massive
compute (since thousands of PDEs are solved) as well as large amounts of
memory (since the adjoint state method requires the forward model to be
saved in memory). Secondly, through the elastic modelling example, we
demonstrate that \href{https://github.com/devitocodes/devito}{Devito}
now fully supports and automates vectorial and second order tensorial
staggered-grid finite-differences with the same high-level API
previously presented for a scalar fields defined on cartesian grids.

This paper is organized as follows: First, we provide a brief overview
of \href{https://github.com/devitocodes/devito}{Devito} and its symbolic
API and present the distributed memory implementation that allows
large-scale modeling and inversion by means of domain decomposition. We
then provide a brief comparison with a state of the art hand-coded wave
propagator to validate the performance previously benchmarked with the
roofline model
(\cite{patterson, devito-compiler, devito-api, louboutin2016ppf}).
Before concluding, results from the Cloud-based experiments discussed
above are presented, highlighting the vectorial and tensorial
capabilities of Devito.

@@

Devito \cite{devito-api} provides a functional language built on top of
\texttt{SymPy} \cite{sympy} to symbolically express PDEs at a
mathematical level and implements automatic discretization with the
finite-difference method. The language is by design flexible enough to
express other types of non finite-diffrence operators, such as
interpolation and tensor contractions, that are inherent to
measurement-based inverse problems. Several additional features are
supported, among which staggered grids, sub-domains, and stencils with
custom coefficients. The last major building block of a good PDE solver
are the boundary conditions, that for finite-difference methods are
notoriously various and often complicated. The system is, however,
sufficiently flexible to express them through composition of core
mechanisms. For example, free surface and perfectly-matched layers
(PMLs) boundary conditions can be expressed as equations -- just like
any other PDE equations -- over a suitable sub-domain.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{devito }\CharTok{import} \NormalTok{Grid}
\NormalTok{grid = Grid(shape=(nx, ny, nz), extent=(ext_x, ext_y, ext_z), origin=(o_x, o_y, o_z))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{devito }\CharTok{import} \NormalTok{Function, TimeFunction}
\NormalTok{m = Function(name=}\StringTok{"m"}\NormalTok{, grid=grid, space_order=so)}
\NormalTok{u = TimeFunction(name=}\StringTok{"u"}\NormalTok{, grid=grid, space_order=so, time_order=to)}
\end{Highlighting}
\end{Shaded}

where \texttt{so} is the spatial discretization order and \texttt{to}
the time discretization order that is used for the generation of the
finite-difference stencil. Second, we define point-wise objects such as
point sources \texttt{src}, located at physical coordinates $x_s$, and
receiver (measurement) objects \texttt{rec}, with sensors located at the
physical locations $x_r$

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{devito }\CharTok{import} \NormalTok{Function, TimeFunction}
\NormalTok{src = SparseFunction(name=}\StringTok{"src"}\NormalTok{, grid=grid, npoint=}\DecValTok{1}\NormalTok{, coordinates=x_s)}
\NormalTok{d = SparseTimeFunction(name=}\StringTok{"d"}\NormalTok{, grid=grid, npoint=}\DecValTok{1}\NormalTok{, nt=nt, coordinates=x_r)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{devito }\CharTok{import} \NormalTok{solve, Eq, Operator}
\NormalTok{eq = m * u.dt2 - u.laplace}
\NormalTok{stencil = [Eq(u.forward, solve(eq, u.forward))]}
\NormalTok{src_eqns = s1.inject(u.forward, expr=s1 * dt**}\DecValTok{2} \NormalTok{/ m)}
\NormalTok{d_eqns = d.interpolate(u)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{devito }\CharTok{import} \NormalTok{Operator}
\NormalTok{op = Operator(stencil + src_eqns + d_eqns)}
\end{Highlighting}
\end{Shaded}

>>>>

\usepackage{listings}
% Define slightly more reasonable Listings defaults
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    prebreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
    frame=lines,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    keywordstyle=\color[gray]{0.4}\bfseries,
    commentstyle=\color[gray]{0.65}\itshape,
    numbers=left,
    captionpos=b,
}

% Default fixed font does not support bold face
\DeclareFixedFont{\ttb}{T1}{txtt}{bx}{n}{9} % for bold
\DeclareFixedFont{\ttm}{T1}{txtt}{m}{n}{9}  % for normal

% Custom colors
\usepackage{color}
\definecolor{deepblue}{rgb}{0,0,0.5}
\definecolor{deepred}{rgb}{0.6,0,0}
\definecolor{deepgreen}{rgb}{0,0.5,0}

% Python style for highlighting
\newcommand\pythonstyle{\lstset{
    language=Python,
    basicstyle=\ttm,
    aboveskip=10pt,
    belowskip=10pt,
    otherkeywords={self},             % Add keywords here
    keywordstyle=\ttb\color{deepblue},
    emph={MyClass,__init__},          % Custom highlighting
    emphstyle=\ttb\color{deepred},    % Custom highlighting style
    stringstyle=\color{deepgreen},
    frame=tb,                         % Any extra options here
    showstringspaces=false            % 
}}

% Python environment
\lstnewenvironment{python}[1][]
{
    \pythonstyle
    \lstset{#1}
}
{}

\usepackage{color}
\usepackage{fancyvrb}
\usepackage{fvextra}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, breaklines=true}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}

@@

It is with these specifications in mind that
\href{https://github.com/devitocodes/devito}{Devito}, a symbolic domain
specific language (DSL) and compiler for the automatic generation of
finite-difference stencils, has been developed. Originally deigned to
accelerate research and development in exploration geophysics, the high-level
interface, previously described in detail in \cite{devito-api}, is built on top
of \texttt{SymPy} \cite{sympy} and is inspired by the underlying mathematical
formulations and other DSLs such as FEniCS \cite{fenics} and Firedrake
\cite{firedrake}. This interface allows the user to formulate wave-equations,
and more generally time-dependent PDEs in a simple and mathematically coherent
way. The Devito compiler then automatically generates finite-difference
stencils from these mathematical expressions. One of the main advantages of
\href{https://github.com/devitocodes/devito}{Devito} over other
finite-difference DSLs is that generic expressions such as sparse operations
(i.e.~point source injection or localized measurements) are fully supported and
expressible in a high-level fashion. The second component of
\href{https://github.com/devitocodes/devito}{Devito} is its compiler (c.f
\cite{devito-compiler}) that generates highly optimized C code. The generated
code is then compiled at runtime for the hardware at hand.

Previous work focused on the DSL and compiler to highlight the potential
application and use cases of Devito. Here, we present a series of
extensions and applications to large-scale three-dimensional problem
sizes as encountered in exploration geophysics. These experiments are
carried out in Cloud-based HPC systems and include elastic forward
modelling using distributed-memory parallelism and imaging based on the
tilted transverse isotropic (TTI) wave-equation
(\cite{virieux, thomsen1986, zhang-tti, duveneck, louboutin2018segeow}).
These proof of concepts highlight two critical features: first, the
ability of the symbolic interface and the compiler to translate to
large-scale adjoint-based inversion problems that require massive
compute (since thousands of PDEs are solved) as well as large amounts of
memory (since the adjoint state method requires the forward model to be
saved in memory). Secondly, through the elastic modelling example, we
demonstrate that \href{https://github.com/devitocodes/devito}{Devito}
now fully supports and automates vectorial and second order tensorial
staggered-grid finite-differences with the same high-level API
previously presented for a scalar fields defined on cartesian grids.

This paper is organized as follows: first, we provide a brief overview
of \href{https://github.com/devitocodes/devito}{Devito} and its symbolic
API and present the distributed memory implementation that allows
large-scale modeling and inversion by means of domain decomposition. We
then provide a brief comparison with a state of the art hand-coded wave
propagator to validate the performance previously benchmarked with the
roofline model
(\cite{patterson, devito-compiler, devito-api, louboutin2016ppf}).
Before concluding, results from the Cloud-based experiments discussed
above are presented, highlighting the vectorial and tensorial
capabilities of Devito.

@@

Devito \cite{devito-api} provides a functional language built on top of
\texttt{SymPy} \cite{sympy} to symbolically express PDEs at a
mathematical level and implements automatic discretization with the
finite-difference method. The language is by design flexible enough to
express other types of non finite-diffrence operators, such as
interpolation and tensor contractions, that are inherent to
measurement-based inverse problems. Several additional features are
supported, among which staggered grids, sub-domains, and stencils with
custom coefficients. The last major building block of a good PDE solver
are the boundary conditions, that for finite-difference methods are
notoriously various and often complicated. The system is, however,
sufficiently general to express them through composition of core
mechanisms. For example, free surface and perfectly-matched layers
(PMLs) boundary conditions can be expressed as equations -- just like
any other PDE equations -- over a suitable sub-domain.

@@

\begin{python}
from devito import Grid
grid = Grid(shape=(nx, ny, nz),
            extent=(ext_x, ext_y, ext_z),
            origin=(o_x, o_y, o_z))
\end{python}

@@

\begin{python}
from devito import Function, TimeFunction
m = Function(name="m", grid=grid,
             space_order=so)
u = TimeFunction(name="u", grid=grid,
                 space_order=so,
                 time_order=to)
\end{python}

where \texttt{so} is the spatial discretization order and \texttt{to}
the time discretization order that is used for the generation of the
finite-difference stencil. Second, we define point-wise objects such as
point sources \texttt{src}, located at physical coordinates $x_s$, and
receiver (measurement) objects \texttt{d}, with sensors located at the
physical locations $x_r$

\begin{python}
from devito import SparseTimeFunction
src = SparseTimeFunction(name="src", grid=grid,
                         npoint=1,
                         coordinates=x_s)
d = SparseTimeFunction(name="d", grid=grid,
                       npoint=1, nt=nt,
                       coordinates=x_r)
\end{python}

@@

\begin{python}
from devito import solve, Eq, Operator
eq = m * u.dt2 - u.laplace
update = Eq(u.forward, solve(eq, u.forward))
src_eqns = s1.inject(u.forward, expr=s1*dt**2/m)
d_eqns = d.interpolate(u)
\end{python}

@@

\begin{python}
from devito import Operator
op = Operator(update + src_eqns + d_eqns)
\end{python}
